Stagnating Public University Finances and Faculty Composition

### 2022-08-09 text/paper.tex
adjusted draft
extra detail.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Title page
% Author
\author{Senan Hogan-Hennessy\thanks{
    Cornell University, Department of Economics.}
    \thanks{I thank Professor Evan Riehl for guidance on this project, Michael Lovenheim, Ronald Ehrenberg, Douglas Miller, and Francine Blau for various comments on the project's direction,
    as well as seminar participants at Cornell University (Spring 2022) for helpful
    discussion.
    This project's Github repository, with materials and (publicly available) data for replication, is available at 
    \url{https://github.com/shoganhennessy/finances-faculty-composition}.
    Any comments or suggestions may be sent to me at \href{mailto:seh325@cornell.edu}{\nolinkurl{seh325@cornell.edu}}, or raised as an issue on the Github project.}}
% Title
\title{Stagnating Public University Finances and Faculty Composition}
\date{\today}
\rhead{\today}
\lhead{Stagnating Public University Finances and Faculty Composition}
% Begin
\begin{document}
\clearpage \maketitle
\thispagestyle{empty}
% Abstract
\begin{abstract}
\noindent
State support for public universities has stagnated the last thirty years, while enrollment has secularly increased.
At the same time, public universities have reduced the number of professors per student on tenure-track or with tenure, while increasing their employment of contingent lecturers.
This analysis relates changes in the faculty composition at public universities to the changes in per student university revenues from state appropriations, addressing endogeneity in state appropriation decisions by a shift-share instrumental variables approach.
An increase of university revenues of 10\%, via a change in state appropriations, increases the number of assistant professors per student by 6.6\% and full professors by 5.3\%, yet decreases the number of lecturers per student by 14.3\%.
Analysis of all the professors hired at Illinois public universities 2011-2021 implies that incumbant professors are not affected by the changes in university revenues, so that changes in faculty composition arises by changes in hiring pratices at public universities.
These results show the compositional effects of stagnating state suport for higher education, and raise questions about the direction that public education is heading thanks to these financial headwinds.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Introduction section
\noindent
Public universities educate the majority of higher education students in the US, yet have experienced a secular decline in state funding (per student) the last few decades.
This decline has been shown to lead to worse education and later-life outcomes for students \citep{NBERw23736,NBERw27885}, yet it is not clear what institutional effects or mechanisms driving effects on students.
I show that four-year, degree-granting public universities (PUs) are exposed to tightening financial constraints thanks to falling state financial support (per student), and show this effects the composition of professors at PUs.
Falls in PU revenues, instrumented by state-level finance shocks, lead to a fall in the number of professors per student within a PU, driven by a fall in assistant or tenured professor counts.
US private universities are not exposed to similar financial constraints during this same time period, and do not exhibit any similar changes in the faculty composition, so that stagnating financial support for PUs have implications for the structure of higher education instruction and research in the US.

Universities in the US are widely considered the highest performing in the world,yet there are consequential differences between its universities that operate in the private sector and those established by state governments.
PUs are subject to numerous state-level administrative laws, and rely on their state governments for funding: an average PU received around \$11,600 per enrolled student in 1990, yet only \$8,300 per enrolled student in 2021.\footnote{
    2021 figures are relatively similar to those for year preceding, despite the Covid-19 pnademic and associated financial shocks, since 2021 refers to the academic year 2019-2021 and so the yearly budget was decided in 2019.
    This paper does not directly analyse budget shocks related to the Covid-19 pandemic, or any other events in 2021 and following.
}
This fall is driven by stagnating funding provided by state governments, while total enrollment among PUs rose by 46\% over the same time period.
At the same time, the number of professors per student at private universities stayed relatively stable at 0.045, yet fell from around 0.04 to around 0.35 professors per student at PUs, driven by a fall of over 20\% in the ratio of associate and full (i.e. tenured) professors per student at PUs.\footnote{
    \autoref{fig:fte-perprof} and \autoref{sec:trends} document these trends between private and public universities in full detail.
}

@@

Findings show that negative changes in university revenues affects faculty composition at public universities, away from tenured and tenure-track professors towards non-tenure track lecturer positions.
A reduction of 10\% in university revenues, via a shock to state appropriations, leads to a fall in 6.6\% in the number of assistant professors per student within a university, 5.3\% for full (tenured) professors, and a reduction in 14.3\% for lecturers.
Additionally, I use individual level data on professors who were hired at Illinois public universities 2011-2021 to investigate whether the shocks to university revenues affected incumbant professors at public universities.
Incumbant professors were not meaningfully affected, in terms of total salary, promotion rate, or rate of leaving the Illinois public university system, which implies that changes in faculty composition came about by changes in the composition of hiring new professors at the university level.

\cite{NBERw23736} first study the effect of (changes in) PUs' finances on student outcomes by isolating changes in PUs' expenditures on student instruction, addressing endogeneity of expenditures decisions by instrumenting expenditures by an instrument for state appropriations and tuition caps.
\cite{NBERw23736} find that increases in PU spending (via state appropriations) increases enrolment and degree completion among students, but corresponding changes in attendance price (via tuition caps/freezes) do not have an effect.
\cite{chakrabarti2018effect,NBERw27885} use the same instrument for PU revenues at the state level, combined with richer information on students' individual-level outcomes, to show that increases in state appropriations lead to lower time to completion and debt levels among students later in life.
\cite{bound2019public} use another variant of the same approach to show that state-level higher education spending cuts induced PUs to shift toward tuition as their primary source of revenue, and shift institutional resources towards students who pay higher tuition rates.
In a similar vein, \cite{bound2007cohort} document variation in per student state funding for higher education resulting from yearly changes in a state's student-age population; they find that lower per student funding lead to lower higher education completion rate among students.

@@

This analysis finds that faculty composition is meaningfully affected by the last few decades' stagnation in state support for higher education.
Analysis of outcomes for incumbant professors in Illinois implies that the explaining mechanism is changes to hiring by public universities in response to the increased financial constraints.
Previous findings from the literature on the effectiveness of tenured and tenure-track faculty versus contingent faculty in the areas of research and teaching.
Combined with this analysis, these findings provide a plausible mechanism for the deteriorating student ouctomes shown to be a result of stagnating finances at public universities \citep{NBERw23736,NBERw27885}.

@@

\begin{table}[h!]
    \onehalfspacing
    \centering
    \caption{IPEDS Summary Statistics, PUs Panel 1987--2021}
    \makebox[\textwidth][c]{\input{tables/ipeds-summary-fte.tex}}
    \label{tab:ipeds-summary}
\end{table}

IPEDS provides information at the university level, yet lacks information on the distribution of salaries and professor count within a university.\footnote{
    IPEDS provides total paid to salary and total faculty employed per university-year, so that I can investigate average professor salary per university-year with IPEDS data.
    Unfortunately, it seems that this measure of professors' salaries is particularly crude in measuring on professors' salaries and other individual-outcomes; summary statistics on IPEDS data do not agree with trends in average professor salary over the sample time period \citep{aau2021survey}.
}
To investigate the distribution, I integregate individual-level data for every PU professor in the state of Illinois between the years 2010-2021.
IBHED host the information;
Public Act 96-0266 (effective 1 January 2010) requires that each PU report to the IBHE the base salary and benefits of the president and all administrators, faculty members, and instructors employed by the college or university.\footnote{
    The universities included are all campuses of the eight Illinois public universities: Chicago State University, Eastern Illinois University, Governors State University, Illinois State University, Northeastern Illinois University, Northern Illinois University, Southern Illinois University  (all five campuses), University of Illinois (all four campuses), Western Illinois University.
}
This publicly available data provides the basis to build a panel of Illinois PU professors 2010-2021; I define a professor as an individual observation by their first plus last name and university pairing, and link this database to IPEDS information on finances for their employing institution.
\begin{table}[h!]
    \onehalfspacing
    \centering
    \caption{IBHED Summary Statistics, Professor Panel 2011--2021.}
    \makebox[\textwidth][c]{\input{tables/illinois-summary-rolling.tex}}
    \label{tab:illinois-summary}
\end{table}
The analysis sample focuses on the subset of professors with observed year of hiring between the years 2011 and 2021, representing 1,778 professors in 2011, and 9,099 in 2021.
\autoref{tab:illinois-summary} presents the summary statistics for this sample.\footnote{
    The full data IBHED contain 173,000 observations for every public university professor in Illinois 2010-2021, which representing 16,932 professors in the year 2010 and 15,352 in the year 2021.
    Summary statistics for the entire sample can be see in Appendix \autoref{tab:illinois-summary-entire}.
    The reasons for restricting to the sample with identified year of hire in the range 2011-2021 is explained in \autoref{sec:iv-model-indiv}.
}

@@

\autoref{fig:funding} exhibits the trends for the mean of US PUs between the years 1990-2021.
We see a rise in total revenues received by PUs (from all sources), and a notable increase in mean tuition revenues from \$48 million per year-university to \$150 million per year-university. 
Non-institutional revenues refer to revenues coming from external sources, not raised by internal university activities;\footnote{
    US universities raise revenues from many different internal activities, including sales of non-degree related educational courses to publishing deals.
    The non-institutional revenue measure is constructed to consider only revenues resulting from external decisions, including state, local, and federal appropriations and tuition revenues.
    Rates of tuition area considered external since most state governments play a large role in deciding tuition rices for PUs \citep{NBERw23736}.
}
non-institutional revenues rose over this time period almost exclusively thanks to the rise in tuition revenues.
At the same time, total state appropriations stagnated at around \$100 million per year-university for 1990-2008, fell slightly around 2008 and have not recovered.
While PUs experienced a financial stagnation, private universities were not exposed to the same constraints, receiving \$37,000 per student in 1990 and \$49,000 in 2021 with no corresponding decline in any specific component.

\begin{figure}[h!]
    \centering
    \caption{Mean Total Revenues among Public Universities, by Year.}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Total, millions \$ 2021 CPI-U.}
        \includegraphics[width=\textwidth]{figures/mean-funding-total.png}
        \label{fig:mean-funding-total}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Per Enrolled Student, \$ 2021 CPI-U.}
        \includegraphics[width=\textwidth]{figures/mean-funding-fte.png}
        \label{fig:mean-funding-fte}
    \end{subfigure}
    \label{fig:funding}
\end{figure}

\begin{figure}[h!]
    \centering
    \caption{Total Student Enrollment, by University Sector, and Year.}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Total.}
        \includegraphics[width=\textwidth]{figures/enrollment-total.png}
        \label{fig:enrollment-total}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Mean, University Level.}
        \includegraphics[width=\textwidth]{figures/enrollment-mean.png}
        \label{fig:enrollment-mean}
    \end{subfigure}
    \label{fig:enrollment}
\end{figure}
At the same time, student enrollment at PUs rose precipitously.
6.2 million students were enrolled in PUs in 1990, and this number rose by 47\% to 9.1 million, with most of the increase occurring over the years 2000-2021.
\autoref{fig:enrollment} shows that total enrollment at private universities has also risen over the same time period, but not as drastic in either relative or absolute terms; the mean public university also grew over this time period, from 9,800 students in 1990 to 11,800 in 2021.
This means that per student mean revenues per student (seen in \autoref{fig:mean-funding-fte}) have stagnated across all measures, and particularly fallen from \$11,000 per student in 1990 to less than \$8,000 per student in 2021.
\begin{figure}[!h]
    \centering
    \caption{Total Professor Count per Student, by University Sector, Professor Appointment, and Year.}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Lecturer}
        \includegraphics[width=\textwidth]{figures/lecturer-fte-perprof.png}
        \label{fig:lecturer-fte-perprof}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Assistant.}
        \includegraphics[width=\textwidth]{figures/assistant-fte-perprof.png}
        \label{fig:assistant-fte-perprof}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Full.}
        \includegraphics[width=\textwidth]{figures/full-fte-perprof.png}
        \label{fig:full-fte-perprof}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{All.}
        \includegraphics[width=\textwidth]{figures/all-fte-perprof.png}
        \label{fig:all-fte-perprof}
    \end{subfigure}
    \label{fig:fte-perprof}
\end{figure}

\autoref{fig:fte-perprof} documents the divergence in faculty composition (per student) between the average private and public university 1990-2021.
Private universities start with  a higher baseline of around 4.5 professors per 100 students, and exhibits yearly variation of less than 0.5 professor per 100 students over the thirty years.
PUs start with 3.9 professors per student, and this number falls to 3.5 primarily in the 2008-2011 time period.
We see a similar difference in baseline, and fall for the years 2008-2011 for assistant professors.
Private and public universities have similar numbers of associated and full professors before the year 2000, yet this number has fallen by over 20\% in the next 20 years only for PUs: in 2021 the mean PU has 6 fewer full professors per student than the mean private university.
At the same time period, we see the rise in use of non-tenure track instructor positions (referred to as lecturers from here on), who were employed at similar rates in both sectors in 1990 yet have been utilised by PUs at a higher rate since.

@@

The state of Illinois particularly exemplifies the stagnation in state support for higher education that the rest of the nation has experienced 1990-2021.
\autoref{fig:illinois-funding} shows the trends for revenues among the seven Illinois public university campuses, including the falling share of state appropriations and substitute towards tuition revenue.

\begin{figure}[h!]
    \centering
    \caption{Mean Revenues among Illinois Public Universities, by Year.}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Total, millions \$ 2021 CPI-U.}
        \includegraphics[width=\textwidth]{figures/illinois-funding-total.png}
        \label{fig:illinois-funding-total}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Per Enrolled Student, \$ 2021 CPI-U.}
        \includegraphics[width=\textwidth]{figures/illinois-funding-fte.png}
        \label{fig:illinois-funding-fte}
    \end{subfigure}
    \label{fig:illinois-funding}
\end{figure}

The decade of 2010-2021 has also seen yearly variation in its revenues, particularly around 2016.
In the calendar year 2015 partisan disagreements between the democratic legislature and republican governor led to the 2016 fiscal year starting with no state budget.
State agencies, and higher education institutions, employed accounting tehniques to continue operation.\footnote{
    Fiscal year 2016 refers to June 205 to June 2016, as is the same for the academic year definition in this analysis. 
    K-12 schools were unaffected thanks to a separate budget agreement allowing non-higher education to operate as before.
}
While state institutions continues to operate, there were drastic revenue and spending cuts in response to the budget impasse, as it continued through fiscal year 2017, and ended with a new budget restoring revenues to state agencies and universities for the 2018 fiscal year.
This variation in public university revenues, stemming entirely from political disagreements --- not from state decisions regarding higher education and its finances \citep{young2020squandered} --- and via the state appropriations channel, mean that Illinois public universities exhibit sizable changes in their state approriations over 2011-2021, of similar order to those for the rest of the country over 1990-2021.

@@

A professor is employed by a university, so it stands to reason that when PUs experience a secular fall in revenue --- analogous to a demand shock in the private sector --- that their human resources decisions will be affected.
Universities have been shown to freeze faculty hiring and general faculty salaries in response to financial shocks, such as the Great Recession \citep{turner2014impact}.

Na\"ively we can analyse the relationship between professor-outcomes $Y_{i,t}$ at university $i$ in year $t$ as a result of university revenues per student $X_{i,t}$.
The exact measure of university revenues, $X_{it}$ is non-institutional revenues, the sume of university funding from outside sources -- local, federal, and state appropriations plus tuition revenue.\footnote{
    This measure is in line with the \cite{NBERw27885} approach of considering revenues from the state, but does not ignore yearly changes in revenues arising from tuition revenues -- which are shown to be heavily related.
}
Outcomes include count of professors per student, and average salary, within each university-year.
This model gives $\beta$ the association with these outcomes and university revenues per student,\footnote{
    Note that dividing by student count also implicitly controls for the size of the university, so that this model accounts for yearly variation in professor count and university revenues arising from growth in a university.
}
including fixed effects to control for effects specific to the university and year.
Log$(.)$ transforming the variables improves the interpretability of $\beta$ as an elasticity for professor count per student with respect to university revenues.
\begin{equation}
    \label{eqn:naivereg}
    Y_{i,t} = \alpha_i + \gamma_t + \beta X_{i,t} + \epsilon_{i,t}
\end{equation}

@@

\cite{NBERw23736,chakrabarti2018effect,NBERw27885} address endogeneity in PU finances by exploiting a shift-share instrument for changes in state-level funding interacted with university reliance on state funding in a base period.
The system exploits the fact that institutions who rely on state appropriations more will be affected by later shocks to state appropriations.
We start with the shift-share instrument.
\begin{align}
    \label{eqn:public-instrument}
    Z_{i,t} &\coloneqq \log \left[
    \left( \frac{\text{Total State Funding}_{s(i),t}}{\text{Student Population}_{s(i),t}} \right)
    \sum_{\tau = 0}^{3} \frac 14
    \left( \frac{\text{State Funding}_{i,1990 + \tau}}{\text{Total Revenues}_{i,1990 + \tau}} \right) \right]
\end{align}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Title page
% Author
\author{Senan Hogan-Hennessy\thanks{
    Cornell University, Department of Economics.}
    \thanks{I thank Professor Evan Riehl for guidance on this project, Michael Lovenheim, Ronald Ehrenberg, Douglas Miller, and Francine Blau for various comments on the project's direction,
    as well as seminar participants at Cornell University (Autumn 202) for helpful
    discussion.
    This project's Github repository, with materials and (publicly available) data for replication, is available at 
    \url{https://github.com/shoganhennessy/finances-faculty-composition}.
    Any comments or suggestions may be sent to me at \href{mailto:seh325@cornell.edu}{\nolinkurl{seh325@cornell.edu}}, or raised as an issue on the Github project.}}
% Title
\title{Stagnating Public University Finances and Faculty Composition}
\date{\today}
\rhead{\today}
\lhead{Stagnating Public University Finances and Faculty Composition}
% Begin
\begin{document}
\clearpage \maketitle
\thispagestyle{empty}
% Abstract
\begin{abstract}
\noindent
State support for public universities has stagnated the last thirty years, while enrollment has secularly increased.
At the same time, public universities have reduced the number of professors per student on tenure-track or with tenure, while increasing their employment of contingent lecturers.
This analysis relates changes in the faculty composition at public universities to the changes in per student university revenues from state appropriations, addressing endogeneity in state appropriation decisions by a shift-share instrumental variables approach.
An increase of university revenues of 10\%, via a change in state appropriations, increases the number of assistant professors per student by 6.6\% and full professors by 5.3\%, yet decreases the number of lecturers per student by 14.3\%.
Analysis of all the professors hired at Illinois public universities 2011-2021 implies that incumbent professors are not affected by the changes in university revenues, so that changes in faculty composition arises by changes in hiring practices at public universities.
These results show the compositional effects of stagnating state support for higher education, and raise questions about the direction that public education heads as these financial headwinds continue.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Introduction section
\noindent
Public universities educate the majority of higher education students in the US, yet have experienced a secular decline in state funding (per student) the last three decades.
This decline has been shown to lead to worse education and later-life outcomes for students \citep{NBERw23736,NBERw27885}, yet it is not clear what mechanisms drive these effects or how public universities are affected as institutions.
I show that four-year, degree-granting public universities have experienced falling state funding, leading to falling revenues per student, effecting the composition of professors within public universities in general.
Falls in public university revenues from the state, instrumented by state-level finance shocks, lead to a fall in the number of tenure-track or tenured professors per student within a university, an increase in the number of lecturers, and overall fall in the number of faculty per student.
US private universities were not exposed to such financial constraints during this same time period, and do not exhibit the substitution away from tenure-track and tenured professors, so that the stagnating state support for public universities has implications for the structure of higher education instruction and research in the US.

US Universities are widely considered the highest performing in the world, yet there are consequential differences between its universities that operate in the private sector and those established by state governments.
public universities are subject to numerous state-level administrative laws, and rely on their state governments for funding: an average PU received around \$11,600 per enrolled student in 1990, yet only \$8,300 per enrolled student in 2021.\footnote{
    2021 figures are relatively similar to those for year preceding, despite the Covid-19 pnademic and associated financial shocks, since 2021 refers to the academic year 2019-2021 and so the yearly budget was decided in 2019.
    This paper does not directly analyse budget shocks related to the Covid-19 pandemic, or any other events in 2021 and following.
}
This fall is driven by stagnating funding provided by state governments, while total enrollment among public universities rose by 46\% over the same time period.
At the same time, the number of professors per student at public universities stayed relatively stable at 0.045, yet fell from around 0.04 to around 0.35 professors per student at public universities, driven by a fall of over 20\% in the ratio of associate and full (i.e. tenured) professors per student at public universities.\footnote{
    \autoref{fig:fte-perprof} and \autoref{sec:trends} document these trends between private and public universities in full detail.
}

@@

Findings show that negative changes in university revenues affects faculty composition at public universities, away from tenured and tenure-track professors towards non-tenure track lecturer positions.
A reduction of 10\% in university revenues, via a shock to state appropriations, leads to a fall in 6.6\% in the number of assistant professors per student within a university, 5.3\% for full (tenured) professors, and a reduction in 14.3\% for lecturers.
Additionally, I use individual level data on professors who were hired at Illinois public universities 2011-2021 to investigate whether the shocks to university revenues affected incumbent professors at public universities.
Incumbent professors were not meaningfully affected, in terms of total salary, promotion rate, or rate of leaving the Illinois public university system, which implies that changes in faculty composition came about by changes in the composition of hiring new professors at the university level.

\cite{NBERw23736} first study the effect of (changes in) public universities' finances on student outcomes by isolating changes in public universities' expenditures on student instruction, addressing endogeneity of expenditures decisions by instrumenting expenditures by an instrument for state appropriations and tuition caps.
\cite{NBERw23736} find that increases in PU spending (via state appropriations) increases enrolment and degree completion among students, but corresponding changes in attendance price (via tuition caps/freezes) do not have an effect.
\cite{chakrabarti2018effect,NBERw27885} use the same instrument for PU revenues at the state level, combined with richer information on students' individual-level outcomes, to show that increases in state appropriations lead to lower time to completion and debt levels among students later in life.
\cite{bound2019public} use another variant of the same approach to show that state-level higher education spending cuts induced public universities to shift toward tuition as their primary source of revenue, and shift institutional resources towards students who pay higher tuition rates.
In a similar vein, \cite{bound2007cohort} document variation in per student state funding for higher education resulting from yearly changes in a state's student-age population; they find that lower per student funding lead to lower higher education completion rate among students.

@@

This analysis finds that faculty composition is meaningfully affected by recent stagnation in state support for higher education.
Analysis of outcomes for incumbant professors in Illinois implies that the explaining mechanism is changes to hiring by public universities in response to the increased financial constraints.
Previous findings from the literature on the effectiveness of tenured and tenure-track faculty versus contingent faculty in the areas of research and teaching.
Combined with this analysis, these findings provide a plausible mechanism for the deteriorating student ouctomes shown to be a result of stagnating finances at public universities \citep{NBERw23736,NBERw27885}.

@@

\begin{table}[h!]
    \singlespacing
    \centering
    \caption{IPEDS Summary Statistics, Public Universities Panel 1987--2021}
    \makebox[\textwidth][c]{\input{tables/ipeds-summary-fte.tex}}
    \label{tab:ipeds-summary}
\end{table}

IPEDS provides information at the university level, yet lacks information on the distribution of salaries and professor count within a university.\footnote{
    IPEDS provides total paid to salary and total faculty employed per university-year, so that I can investigate average professor salary per university-year with IPEDS data.
    Unfortunately, it seems that this measure of professors' salaries is particularly crude in measuring on professors' salaries and other individual-outcomes; summary statistics on IPEDS data do not agree with trends in average professor salary over the sample time period \citep{aau2021survey}.
}
To investigate the distribution, I integregate individual-level data for every PU professor in the state of Illinois between the years 2010-2021.
IBHED host the information;
Public Act 96-0266 (effective 1 January 2010) requires that each university report base salary and benefits all administrators, faculty members, and instructors employed by the college or university.\footnote{
    The universities included are all campuses of the eight Illinois public universities: Chicago State University, Eastern Illinois University, Governors State University, Illinois State University, Northeastern Illinois University, Northern Illinois University, Southern Illinois University  (all five campuses), University of Illinois (all four campuses), Western Illinois University.
}
This publicly available data provides the basis to build a panel of Illinois PU professors 2010-2021; I define a professor as an individual observation by their first plus last name and university pairing, and link this database to IPEDS information on finances for their employing institution.

The analysis sample represents 16,932 professors in the year 2010 and 15,352 in the year 2021, with summary statistics presented in \autoref{tab:illinois-summary}.
Additionally, further analysis using the first year of a professor's employment focuses on the subset of professors with observed year of hiring between the years 2011 and 2021, representing 1,778 professors in 2011, and 9,099 in 2021.\footnote{
    The reasons for restricting to the sample with identified year of hire in the range 2011-2021 is explained in \autoref{sec:iv-model-indiv}.
}
Summary statistics for this group, which over-represents lecturers compared to assistant and full professors, can be seen in \autoref{tab:illinois-summary-rolling}.

\begin{table}[h!]
    \singlespacing
    \centering
    \caption{IBHED Summary Statistics, Professor Panel 2010--2021.}
    \makebox[\textwidth][c]{\input{tables/illinois-summary.tex}}
    \label{tab:illinois-summary}
\end{table}

@@

\autoref{fig:funding} exhibits the trends for the mean of US public universities between the years 1990-2021.
We see a rise in total revenues received by public universities (from all sources), and a notable increase in mean tuition revenues from \$48 million per year-university to \$150 million per year-university. 
Non-institutional revenues refer to revenues coming from external sources, not raised by internal university activities;\footnote{
    US universities raise revenues from many internal activities, including sales of nondegree related educational courses to publishing deals.
    The non-institutional revenue measure is constructed to consider only revenues resulting from external decisions, including state, local, and federal appropriations and tuition revenues.
    Rates of tuition are considered external since most state governments play a large role in deciding tuition rices for public universities \citep{NBERw23736}.
}
non-institutional revenues rose over this time period almost exclusively thanks to the rise in tuition revenues.
At the same time, total state appropriations stagnated at around \$100 million per year-university for 1990-2008, fell slightly around 2008 and have not recovered.
While public universities experienced a financial stagnation, private universities were not exposed to the same constraints, receiving \$37,000 per student in 1990 and \$49,000 in 2021 with no corresponding decline in any specific component.

\begin{figure}[h!]
    \centering
    \singlespacing
    \caption{Mean Total Revenues among Public Universities, by Year.}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Total, millions \$ 2021 CPI-U.}
        \includegraphics[width=\textwidth]{figures/mean-funding-total.png}
        \label{fig:mean-funding-total}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Per Enrolled Student, \$ 2021 CPI-U.}
        \includegraphics[width=\textwidth]{figures/mean-funding-fte.png}
        \label{fig:mean-funding-fte}
    \end{subfigure}
    \label{fig:funding}
\end{figure}

\begin{figure}[h!]
    \centering
    \singlespacing
    \caption{Total Student Enrollment, by University Sector, and Year.}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Total.}
        \includegraphics[width=\textwidth]{figures/enrollment-total.png}
        \label{fig:enrollment-total}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Mean, University Level.}
        \includegraphics[width=\textwidth]{figures/enrollment-mean.png}
        \label{fig:enrollment-mean}
    \end{subfigure}
    \label{fig:enrollment}
\end{figure}
At the same time, student enrollment at public universities rose precipitously.
6.2 million students were enrolled in public universities in 1990, and this number rose by 47\% to 9.1 million, with most of the increase occurring over the years 2000-2021.
\autoref{fig:enrollment} shows that total enrollment at public universities has also risen over the same time period, but not as drastic in either relative or absolute terms; the mean public university also grew over this time period, from 9,800 students in 1990 to 11,800 in 2021.
This means that per student mean revenues per student (seen in \autoref{fig:mean-funding-fte}) have stagnated across all measures, and particularly fallen from \$11,000 per student in 1990 to less than \$8,000 per student in 2021.
\begin{figure}[!h]
    \centering
    \singlespacing
    \caption{Total Professor Count per Student, by University Sector, Professor Appointment, and Year.}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Lecturer}
        \includegraphics[width=\textwidth]{figures/lecturer-fte-perprof.png}
        \label{fig:lecturer-fte-perprof}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Assistant.}
        \includegraphics[width=\textwidth]{figures/assistant-fte-perprof.png}
        \label{fig:assistant-fte-perprof}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Full.}
        \includegraphics[width=\textwidth]{figures/full-fte-perprof.png}
        \label{fig:full-fte-perprof}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{All.}
        \includegraphics[width=\textwidth]{figures/all-fte-perprof.png}
        \label{fig:all-fte-perprof}
    \end{subfigure}
    \label{fig:fte-perprof}
\end{figure}

\autoref{fig:fte-perprof} documents the divergence in faculty composition (per student) between the average private and public university 1990-2021.
Public universities start with  a higher baseline of around 4.5 professors per 100 students, and exhibits yearly variation of less than 0.5 professor per 100 students over the thirty years.
Public universities start with 3.9 professors per student, and this number falls to 3.5 primarily in the 2008-2011 time period.
We see a similar difference in baseline, and fall for the years 2008-2011 for assistant professors.
Private and public universities have similar numbers of associated and full professors before the year 2000, yet this number has fallen by over 20\% in the next 20 years only for public universities: in 2021 the mean PU has 6 fewer full professors per student than the mean private university.
At the same time period, we see the rise in use of non-tenure track instructor positions (referred to as lecturers from here on), who were employed at similar rates in both sectors in 1990 yet have been utilised by public universities at a higher rate since.

@@

The state of Illinois particularly exemplifies the stagnation in state support for higher education that the rest of the nation has experienced 1990-2021.
\autoref{fig:illinois-funding} shows the trends for revenues among the seven Illinois public university campuses, including the falling share of state appropriations and substitution towards tuition revenue.

\begin{figure}[h!]
    \centering
    \singlespacing
    \caption{Mean Revenues among Illinois Public Universities, by Year.}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Total, millions \$ 2021 CPI-U.}
        \includegraphics[width=\textwidth]{figures/illinois-funding-total.png}
        \label{fig:illinois-funding-total}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Per Enrolled Student, \$ 2021 CPI-U.}
        \includegraphics[width=\textwidth]{figures/illinois-funding-fte.png}
        \label{fig:illinois-funding-fte}
    \end{subfigure}
    \label{fig:illinois-funding}
\end{figure}

The decade of 2010-2021 has also seen yearly variation in its revenues, particularly around 2016.
In the calendar year 2015 partisan disagreements between the democratic legislature and republican governor led to the 2016 fiscal year starting with no state budget.
State agencies, and higher education institutions, employed accounting tehniques to continue operation.\footnote{
    Fiscal year 2016 refers to June 205 to June 2016, as is the same for the academic year definition in this analysis. 
    K-12 schools were unaffected thanks to a separate budget agreement allowing non-higher education to operate as before.
}
While state institutions continued to operate, there were drastic revenue and spending cuts in response to the budget impasse, as it continued through fiscal year 2017, and ended with a new budget restoring revenues to state agencies and universities for the 2018 fiscal year.
This variation in public university revenues, stemming entirely from political disagreements --- not from state decisions regarding higher education and its finances \citep{young2020squandered} --- via the state appropriations channel, mean that Illinois public universities exhibit sizable changes in their state appropriations over 2011-2021, of similar order to those for the rest of the country over 1990-2021.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Section on possible outcomes
\subsection{University Responses to Financial Shocks}
\label{sec:responses}
Subsection describing how a university may respond to shocks on its sources of revenues.
Cite other papers on how this went.

Write about shocks in state appropriations primarily here.

Additional paragraph on how the incumbent professors may be affected.

@@

A professor is employed by a university, so it stands to reason that when public universities experience a secular fall in revenue --- analogous to a demand shock in the private sector --- that their human resources decisions will be affected.
Universities have been shown to freeze faculty hiring and general faculty salaries in response to financial shocks, such as the Great Recession \citep{turner2014impact}.

\textbf{Shock to state appropriations only, as shocks to total revenues/spending requires identification for multiple sources.
Tuition is an outcome of changes in the state appropriations, so discuss whether it is an outcome or a control.
Leave footnote that results are relatively similar, so leave out this control.}

Na\"ively we can analyse the relationship between professor-outcomes $Y_{i,t}$ at university $i$ in year $t$ as a result of university revenues per student $X_{i,t}$.
The exact measure of university revenues, $X_{it}$ is non-institutional revenues, the sum of university funding from outside sources -- local, federal, and state appropriations plus tuition revenue.\footnote{
    This measure is in line with the \cite{NBERw27885} approach of considering revenues from the state, but does not ignore yearly changes in revenues arising from tuition revenues -- which are shown to be heavily related.
}
Outcomes include count of professors per student, and average salary, within each university-year.
This model gives $\beta$ the association with these outcomes and university revenues per student,\footnote{
    Note that dividing by student count also implicitly controls for the size of the university, so that this model accounts for yearly variation in professor count and university revenues arising from growth in a university.
}
including fixed effects to control for effects specific to the university and year.
Log$(.)$ transforming the variables improves the interpretability of $\beta$ as an elasticity for professor count per student with respect to university revenues.
\begin{equation}
    \label{eqn:naivereg}
    Y_{i,t} = \alpha_i + \gamma_t + \beta X_{i,t} + \epsilon_{i,t}
\end{equation}

@@

\cite{NBERw23736,chakrabarti2018effect,NBERw27885} address endogeneity in PU finances by exploiting a shift-share instrument for changes in state-level funding interacted with university reliance on state funding in a base period.
The system exploits the fact that institutions who rely on state appropriations more will be affected by state appropriation shocks.
We start with the shift-share instrument.
\begin{align}
    \label{eqn:public-instrument}
    Z_{i,t} &\coloneqq \log \left[
    \left( \frac{\text{Total State Funding}_{s(i),t}}{\text{Student Population}_{s(i),t}} \right)
    \sum_{\tau = 0}^{3} \frac 14
    \left( \frac{\text{State Funding}_{i,1990 + \tau}}{\text{Total Revenues}_{i,1990 + \tau}} \right) \right]
\end{align}
